Modeling the Java Card API in B \thanks{This work is partly supported by CAPES and CNPq grants 306033/2009-7, 560014/2010-4 and 573964/2008-4 (National Institute of Science and Technology for Software Engineering---INES, \texttt{www.ines.org.br}).}

### 2012-04-20 trunk/papers/B2012-work/javacardAPIpaper.tex
correcoes de david

In the context of the BSmart project, we are interested in the use of
B~\cite{BBook} to develop Java Card~\cite{chen:2000} applications.  Java Card is a
restricted and optimized version of Java which allows memory and
processor constrained devices, such as smart cards, to store and run
small applications.  Java Card applications have characteristics
(small size and need for security) which make them good candidates for
formal development with the B method. They also present some
regularity of structure that led our group to define a method and
implement a tool which specialize B for the development of such
applications, providing greater support for the developer than pure B
tools \cite{atelierB,proB}. They are the BSmart method and tool,
presented in \cite{BSmart08,Gomes10}.

Previous works have also dealt with B for Smart Cards,
\cite{Bert:2003,JBtools,tatibouet:lncs,requetBtoC}, but none has, at
our knowledge, specified Java Card in B so that the formal development
includes Java Card aspects. In the first versions of BSmart
\cite{Gomes10}, these aspects were included by the code generator, and
in JBTools \cite{JBtools}, the generated code needed to be edited by
the developer to include these aspects. These works lacked then the ability to formally verify the Java Card
dependent aspects of the application. To be able to change this, a B
specification of these aspects is needed, and this is the contribution
of this paper: the B specification of the Java Card API, with its
primitive types and all the classes of the 14 packages composing the
API. This specification can be used with the BSmart tool, 
independently, in any B tool, or even as a complement to the official
Java Card documentation \cite{oracle}.

In the following we present some important aspects of the Java Card platform, in section \ref{sec:javacard}, and we introduce the B model of the Java Card API,  in section \ref{sec:api-spec}. We conclude with some notes on the use of the proposed model, in section \ref{sec:using-the-API}, and some conclusions and future work (in section \ref{sec:conc}).

@@

The Java Card API is the part of the Java Card platform that a developer of any Java Card application must use. It defines the constituents of an applet, provides the means to deal with the APDU protocol and communicate data and commands between the host application and the card, and defines some basic services for security and cryptography implementation.

@@

The change of interface is also taken care of by the BSmart method, so
that the development can be done without need to extend the B
method, as presented in~\cite{Gomes10}.

@@

\item Only the abstract models corresponding to each class or
  interface have been developed. Two reasons led to this choice: (1)
  code is not to be generated from the specification of the API, as we
  are only creating an abstract layer to existing implementations; and
  (2) in B, only the abstract machines are seen by modules that build
  upon others, through the SEES, USES, IMPORTS, INCLUDES and EXTENDS
  clauses.

@@

  \item In the current version of the specification, states are not modeled. 
  The B machines corresponding to Java Card classes or interfaces are stateless machines.

@@

\item The hierarchy of the Java Card API objects is partially modeled
  by abstract set inclusion. For instance, the class Throwable extends
  the class Object, in Java Card. In our model, we  have that the
  machine Throwable SEES the machine Object, and to show that all throwables are indeed
  objects, a property stating that THROWABLE is a subset of OBJECT is
  included.

@@

A small piece of specification with some of the characteristics described above is presented in Listing \ref{lis:object}. It presents two \texttt{java.lang} classes: \texttt{Object} and \texttt{Throwable}.  \texttt{Object} is  superclass to  all API classes, with its only method  \texttt{equals} that checks if an object (from which a reference is received as argument) is the same as the current one. \texttt{Throwable} extends \texttt{Object}, but does not define any other method nor redefines \texttt{equals}. The inheritance hierarchy is partially simulated in B by the corresponding set inclusion. These abstract sets are then used to refer to the corresponding class as a type as can be seen in Listing \ref{lis:class-types}.

@@

One delicate issue to be dealt with both by the user of the API
specification and the tools designed to work with it, is the use of
overloaded methods in an application development. When overloaded
operations are used in an application development, the user must
explicitly choose which version (in B) of the overloaded operation is to be
used. Then, verifications will be carried out considering this
specific version of the operation. When code is generated, either the
code generator will remove the name extension automatically, if
designed for it, or the user will have to edit the generated code
manually. In any case, it is important that the actual method called
in the Java Card code is the one with which verifications have been
carried out.  In case of overriding, in particular, it will be either
the user's responsibility to choose the B one which is in scope at
this specific point in Java, or an extra verification step will be
included in the code generator to check that the operation used in
B is the one in scope in Java and registers an error in case it is
not.

@@

Among the limitations of this work, some may be dealt with by adapted tools and some will actually require some special attention from the developer, even with tool support, such as method overloading and overriding . These are the characteristics that are related to the fact that B is not object oriented and does not include exception raising and handling mechanisms. Some work on this last item can be seen on \cite{BR2003}, but it has not been implemented.

With the current level of abstraction of the model and the fact that the Java Card Runtime Environment (JCRE) has not been modeled, different conditions which may lead to exception raising are not yet modeled. Being able to model them would make developed applications even more sure, as more static verification could be carried
out. Continuation of this work then will enlarge the scope of the modeling to cope statically with a larger set of situations.

>>>>

In the context of the BSmart project, we are interested in the use of
B~\cite{BBook} to develop Java Card~\cite{chen:2000} applications.  Java Card is a
restricted and optimized version of Java which allows memory and
processor constrained devices, such as smart cards, to store and run
small applications.  Java Card applications have characteristics
(small size and need for security) which make them good candidates for
formal development with the B method. They also present some
regularity of structure that led our group to define a method and
implement a tool which specializes B for the development of such
applications, providing more focussed support for the developer than pure B
tools \cite{atelierB,proB}. They are the BSmart method and tool,
presented in \cite{BSmart08,Gomes10}.

Previous works have also dealt with B for Smart Cards,
\cite{Bert:2003,JBtools,tatibouet:lncs,requetBtoC}, but none has, to
our knowledge, specified Java Card in B so that the formal development
includes Java Card aspects. In the first versions of BSmart
\cite{Gomes10}, these aspects were included by the code generator. In 
contrast, with the JBTools~\cite{JBtools} approach, the generated code needed to be edited by
the developer to include these aspects. These works lacked then the ability to formally verify the Java Card
dependent aspects of the application. To change this, a B
specification of these aspects is needed, and this is the contribution
of this paper: the B specification of the Java Card API, with its
primitive types and all the classes of the 14 packages composing the
API. This specification can be used with the BSmart tool, 
independently, in any B tool, or even as a complement to the official
Java Card documentation \cite{oracle}.

In the following, we present some important aspects of the Java Card platform, in section \ref{sec:javacard}, and we introduce the B model of the Java Card API,  in section \ref{sec:api-spec}. We conclude with some notes on the use of the proposed model, in section \ref{sec:using-the-API}, and some conclusions and future work (in section \ref{sec:conc}).

@@

The Java Card API is the part of the Java Card platform that a developer of any Java Card application must use. It defines the constituents of an applet, provides the means to deal with the APDU protocol and to communicate data and commands between the host application and the card. It also defines some basic security services such as encryption.

@@

The change of interface is also taken care of by the BSmart method, so
that the development can be done without extending the B
method, as presented in~\cite{Gomes10}.

@@

\item Only the abstract models corresponding to each class or
  interface have been developed. Two reasons led to this choice: (1)
  code is not to be generated from the specification of the API, as we
  are only creating a model fo existing implementations; and
  (2) in B, only the abstract machines are seen by modules that build
  upon others, through the SEES, USES, IMPORTS, INCLUDES and EXTENDS
  clauses.

@@

\item In the current version of the specification, states are not
  modeled.  The B machines corresponding to Java Card classes or
  interfaces are stateless.

@@

\item The hierarchy of the Java Card API objects is partially modeled
  by abstract set inclusion. For instance, the class Throwable extends
  the class Object, in Java Card. In our model, we  have that the
  machine Throwable SEES the machine Object, and to constrain that all throwables are indeed
  objects, a property stating that THROWABLE is a subset of OBJECT is
  included.

@@

A small piece of specification with some of the characteristics described above is presented in Listing \ref{lis:object}. It presents two \texttt{java.lang} classes: \texttt{Object} and \texttt{Throwable}.  \texttt{Object} is a superclass to  all API classes, with a single method  \texttt{equals} that checks if an object (from which a reference is received as argument) is the same as the current one. \texttt{Throwable} extends \texttt{Object}, but does not define any other method nor redefines \texttt{equals}. The inheritance hierarchy is partially simulated in B by the corresponding set inclusion. These abstract sets are then used to refer to the corresponding class as a type as can be seen in Listing \ref{lis:class-types}.

@@

One subtle issue to be dealt with both by the user of the API
specification and the tools designed to work with it, is the use of
overloaded methods in an application development. When overloaded
operations are used in an application development, the user must
explicitly choose which version (in B) of the overloaded operation is to be
used. Then, verifications will be carried out considering this
specific version of the operation. When code is generated, either the
code generator will remove the name extension automatically, if
designed for it, or the user will have to edit the generated code
manually. In any case, it is important that the actual method called
in the Java Card code is the one with which verifications have been
carried out.  In case of overriding, in particular, it will be either
the user's responsibility to choose the B operation in scope at
this specific point in Java, or an extra verification step will be
included in the code generator to check that the operation used in
B is the one in scope in Java and registers an error in case it is
not.

@@

Among the limitations of this work, some may be dealt with by adapted tools and some will actually require some special attention from the developer, even with tool support, such as method overloading and overriding . These are consequence of B not being object oriented and not providing exception raising and handling mechanisms. Some work on this last item can be seen on \cite{BR2003}, but it has not been implemented.

With the current level of abstraction of the model and the fact that the Java Card Runtime Environment (JCRE) has not been modeled, different conditions which may lead to exception raising are not yet modeled. Extending the model to include them would make developed applications more reliable, as more static verification could be carried
out. Continuation of this work then will enlarge the scope of the modeling to cope statically with a larger set of situations.
